Effects of Different Pain Relievers and Dosages on Cognitive Retention

### 2025-03-03 paper/EDA.qmd
Latest Changes, with assumptions explained

## Study Design  
This study involves *7 treatment groups*, with **30 participants** assigned to each. The groups are as follows:

1. Aspirin 500 mg
2. Aspirin 1000 mg
3. Paracetamol 500 mg
4. Paracetamol 1000 mg
5. Tramadol 50 mg
6. Tramadol 100 mg
7. Placebo

This results in **210 total observations** (7 × 30 = 210).

@@

## Data Analysis Method   
We will use **one-way ANOVA** to analyze the effects of pain reliever type on memory performance.
and dosage levels on cognitive retention.

@@

@tbl-summary-2 shows the distribution of participants across drug types. Aspirin, Paracetamol, and Tramadol each have 60 participants (28.57\%), ensuring balance. However, the Placebo group is smaller (30 participants, 14.29\%).

@tbl-summary-3 shows the dosage levels assigned to participants. There is an equal split between high-dose (90 participants, 50\%) and low-dose (90 participants, 50\%) conditions.

@tbl-summary-4 provides an overview of participant locations. The distribution is relatively even across six different locations, with each site having either 30 (14.29\%) or 60 (28.57\%) participants.

@@

@tbl-summary-5 shows the breakdown of age groups, which is a confounding variable in this study. The three age groups: 18–34, 35–50, and 50+ each have 70 participants (33.33\%).

@@

```{r}
#| label: tbl-summary-4
#| tbl-cap: "Summary of Locations"
#| echo: false
#| warning: false
#| message: false

# Count and percentage for Location
location_counts <- table(clean_data$location)
location_percentages <- prop.table(location_counts) * 100
location_summary <- data.frame(Location = names(location_counts), Count = as.numeric(location_counts), Percentage = round(as.numeric(location_percentages), 2))

kable(location_summary, format = "latex", booktabs = TRUE)

@@

1. The median scores for all drugs appear similar before treatment.
2. **Tramadol has a higher median** than other drugs both before and after treatment.
3. The **Placebo group has a slightly lower median score** compared to other drugs.
4. There is **one outlier in the Tramadol group before treatment**.
5. The (IQR) is **similar across all drug types**.

@@

1. Before treatment, the scores are relatively high across all groups, with Paracetamol and Placebo showing slightly higher medians than the others.
2. After treatment, the distributions remain similar, though there is a slight increase in spread  for Aspirin and Paracetamol.
3. Outliers are present in all groups, indicating some variability in memory card performance across individuals.

@@

```{r}
#| echo: false
#| warning: false
#| message: false
#| fig.width: 7
#| fig.height: 4
#| dpi: 300
#| fig-cap: "Distribution of Dosage Levels"

ggplot(clean_data, aes(x = dosage_level, fill = dosage_level)) +
  geom_bar() +
  labs(title = "Distribution of Dosage Levels", x = "Dosage Level", y = "Count") +
  theme_minimal()

```{r}
#| echo: false
#| warning: false
#| message: false
#| fig.width: 6
#| fig.height: 4
#| fig-cap: "Time Taken for Memory Game after taking Drugs"

# Boxplot to visualize the distribution of time taken to solve Memory game after taking pain killers
ggplot(clean_data, aes(x = drug, y = game_after, fill = dosage_level)) +
  geom_boxplot() +
  ggtitle("Time Taken for Memory Game after taking Drugs")

@@

```{r}
#| include: false
#| warning: false
#| message: false
# Barlett Test Homegenity of Variance: testing for ass
# bartlett_test <- bartlett.test(game_after ~ drug, data = clean_data)
# print(bartlett_test)

```{r}
#| echo: false
#| warning: false
#| message: false
#| fig.width: 6
#| fig.height: 4
#| fig-cap: "Time Taken for Memory Game after taking Drugs"

@@

summary(model_1)
```

```{r}
#| echo: false
#| warning: false
#| message: false
# bartlett_test_part1 <- bartlett.test(game_after ~ drug, data = clean_data)
# print(bartlett_test_part1)
```

```{r}
#| echo: false
#| warning: false
#| message: false
#| fig.width: 6
#| fig.height: 4
#| fig-cap: "Residual Plot Model 1"

# QQ plot for Normality assumption
ggqqplot(residuals(model_1)) +
  ggtitle("QQ Plot for Residuals Model 1")

# Shapiro-Wilk Tet
# shapiro_test <- shapiro.test(residuals(model_1))
# print(shapiro_test)

@@

```{r}
#| echo: false
#| warning: false
#| message: false
#| fig.width: 6
#| fig.height: 4
#| fig-cap: "Normality & Homegenity"

@@

# Shapiro-Wilk Test for Normality of Residuals
# shapiro_test_part2 <- shapiro.test(residuals(model_part2))
# print(shapiro_test_part2)

```{r}
#| echo: false
#| warning: false
#| message: false
#| fig.width: 6
#| fig.height: 4
#| fig-cap: "Residual Plot Model"

# QQ plot for residuals of Part 2
ggqqplot(residuals(model_2)) +
  ggtitle("QQ Plot Two-Way ANOVA Model")

# References

>>>>

## Study Design  
This study involves *8 treatment groups*, with **30 participants** assigned to each. The groups are as follows:

1. Aspirin 500 mg (Low)
2. Aspirin 1000 mg (High)
3. Paracetamol 500 mg (Low)
4. Paracetamol 1000 mg (High)
5. Tramadol 50 mg (Low)
6. Tramadol 100 mg (High)
7. Placebo (Low) 
8. Placebo (High)

This results in **240 total observations** (8 × 30 = 240).

@@

## Quantitative and Qualitative Variables 
In our study the quantitative and qualitative variables are as follows. 
Quantitative variables: Memory Game Scores, Memory Test Cards Scores 
Qualitative variables: Type of pain reliever, dosage level (low/high)

## Data Analysis Method   
We will use **one-way ANOVA** to analyze the effects of different pain relievers on memory performance
Similarly, we will use **two-way ANOVA** to analyze the effects of different dosage levels per drug type on memory performance and cognitive retention.

@@

@tbl-summary-2 shows the distribution of participants across drug types. Aspirin, Paracetamol, and Tramadol, and Placebo each have 60 participants (25\%), ensuring balance.

@tbl-summary-3 shows the dosage levels assigned to participants. There is an equal split between high-dose (120 participants, 50\%) and low-dose (120 participants, 50\%) conditions.

@@

@tbl-summary-5 shows the breakdown of age groups, which is a confounding variable in this study. The three age groups: 18–34, 35–50, and 50+ each have 80 participants (33.33\%).

@@

@@

1. The median scores for all drugs appear similar before treatment.
2. **Tramadol has a higher median** than other drugs both before and after treatment.
3. The **Placebo group has a slightly lower median score** compared to other drugs.
4. There is **one outlier in the Tramadol group before treatment**.
5. The (IQR) is **similar across all drug types**, though Tramadol sees a decrease in its IQR in the memory game scores after treatment.

@@

1. Before treatment, the scores are relatively high across all groups, with Paracetamol and Placebo showing slightly higher medians than the others.
2. After treatment, the distributions remain similar, though there is a slight increase in spread for Aspirin, Paracetamol, and Placebo.
3. There is a slight decrease in the IQR of Tramadol. 
3. Outliers are present in all groups, indicating some variability in memory card performance across individuals.

@@

@@

## Checking Model Assumptions

We assessed the homogeneity of variance and normality of residuals using diagnostic plots to validate the assumptions of our ANOVA tests. (@fig-assump-one & @fig-assump-two).

**One-way ANOVA**

*Independence* For our one-way ANOVA model, we have independence because our observations are independent within and across the groups i.e all participants were randomly assigned a pain reliever, and there was no relationships between observations (i.e repeated  participants, treatments etc.)

*Homogeneity of Variance:* 
The reference line in @fig-assump-one relatively flat and horizontal which indicates that variance is approximately equal across groups. Thus, the assumption of homogeneity of variance is not significantly violated.

*Normality of Residuals:* 
The residuals are approximately normally distributed in @fig-assump-one since the density plot on the right appears close to a normal distribution.

**Two-way ANOVA**

*Independence*
For our two-way ANOVA model, the different combinations of pain reliever type and dosage levels were randomly assigned to all participants, in random order.

*Homogeneity of Variance:* 
In @fig-assump-two the left plot shows some deviation from a perfectly flat line and has minor fluctuations, especially at certain levels of drug type and dosage. Despite some variation, the overall trend does not show extreme violations of homogeneity.

*Normality of Residuals:* 
In @fig-assump-two the density plot on the right shows that the residuals follow a roughly normal distribution even though minor deviations are present.

**Conclusion**

Both models satisfy the assumptions of homogeneity of variance and normality of residuals well enough. There are slight deviations in the two-way ANOVA model, but they are not severe enough to invalidate the analysis. Therefore, we proceed with the ANOVA tests.

```{r}
#| label: fig-assump-one
#| echo: false
#| warning: false
#| message: false
#| fig.width: 6
#| fig.height: 4
#| fig-cap: "Homogeneity and Normality (One-way ANOVA)"

@@

@@

```{r}
#| label: fig-assump-two
#| echo: false
#| warning: false
#| message: false
#| fig.width: 6
#| fig.height: 4
#| fig-cap: "Normality & Homegenity (Two-way ANOVA)"

@@
